Sparse Function-space Representation \\ of Neural Networks

### 2023-05-11 paper/main.tex
Update fig
rewritten some equations in the method

@@

\newcommand{\vm}{\mbf{m}}
%\newcommand{\vz}{\mbf{z}}
%\newcommand{\vf}{\mbf{f}}
\newcommand{\vu}{\mbf{u}}
\newcommand{\vx}{\mbf{x}}
\newcommand{\vy}{\mbf{y}}

@@

@@

\section{Introduction}
\label{sec:intro}
%

\subsection{Related Work}
\label{sec:related}
%

% Introduce the laplace GNN + shortcomings
One common approach for uncertainty with neural networks is a Laplace GNN approximation \citep{daxbergerLaplace2021}. The idea is to essentially takes a trained neural network and linearises around the optimal weights. The resulting linear model can be used to obtain uncertainty estimates and sometimes improved predictions \citep{immerScalable2021}. Given the resultant model is now linear with respect to the weights its is also now possible to convert to a Gaussian process as shown in \citep{immerScalable2021, khan2019approximate}. Yet to obtain uncertainty estimates there is the usual computational scaling of the GP of $O(n^3)$, so cubic in the number of data points. Many approaches  apply crude approximations and yet still obtain good uncertainty estimates.

Deep learning has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations when applied to sequential learning paradigms, such as Continual Learning (CL, \cite{add}) and Reinforcement Learning (RL, \cite{add}). In CL, the challenge is retaining a compact representation of the problem and avoid forgetting over the life-long learning horizon (see, \eg, \cite{add}). Similarly in RL, the model would need to adapt to observations of the environment through exploration, but this can be inefficient and would require constant retraining (\eg, \cite{add, add}).

@@

@@

\begin{figure}[t!]
%  \begin{subfigure}{.48\textwidth}
%    \centering
%    \tikz\node[fill=black!10,minimum width=\figurewidth,minimum height=\figureheight,rounded corners=5pt]{};
%  \end{subfigure}
%  \hfill
%  \begin{subfigure}{.48\textwidth}
%    \centering
%    \tikz\node[fill=black!10,minimum width=\figurewidth,minimum height=\figureheight,rounded corners=5pt]{};
%  \end{subfigure}

  \centering\scriptsize
  % Figure options
  \pgfplotsset{axis on top,scale only axis,width=\figurewidth,height=\figureheight, ylabel near ticks,ylabel style={yshift=-2pt},y tick label style={rotate=90},legend style={nodes={scale=1., transform shape}},tick label style={font=\tiny,scale=1}}
  \pgfplotsset{xlabel={Input, $x$},axis line style={rounded corners=3pt}}
  % Set figure 
  \setlength{\figurewidth}{.28\textwidth}
  \setlength{\figureheight}{\figurewidth}
  %
  \def\inducing{\normalsize Sparse inducing points}
  %
  \begin{subfigure}[c]{.34\textwidth}
    \raggedleft
    \pgfplotsset{ylabel={Output, $y$}}
    \input{./fig/regression-nn.tex}%
  \end{subfigure}
  \hfill  
  \begin{subfigure}[c]{.01\textwidth}
    \centering
    \tikz[overlay,remember picture]\node(p0){.};
  \end{subfigure}  
  \hfill
  \begin{subfigure}[c]{.28\textwidth}
    \raggedleft
    \pgfplotsset{yticklabels={,,},ytick={\empty}}
    \input{./fig/regression-nn2svgp.tex}%
  \end{subfigure}
  \hfill  
  \begin{subfigure}[c]{.01\textwidth}
    \centering
    \tikz[overlay,remember picture]\node(p1){.};
  \end{subfigure}  
  \hfill
  \begin{subfigure}[c]{.28\textwidth}
    \raggedleft
    \pgfplotsset{yticklabels={,,},ytick={\empty}}        
    \input{./fig/regression-update.tex}%
  \end{subfigure}
  \caption{\textbf{Regression example on an MLP with two hidden layers.} Left:~Predictions from the trained neural network. Center:~Our approach summarizes the...}
  \label{fig:teaser} 
  % 
  \begin{tikzpicture}[remember picture,overlay]
    % Arrow style
    \tikzstyle{myarrow} = [draw=black!80, single arrow, minimum height=14mm, minimum width=2pt, single arrow head extend=4pt, fill=black!80, anchor=center, rotate=0, inner sep=5pt, rounded corners=1pt]
    % Arrows
    \node[myarrow] (p0-arr) at ($(p0) + (0em,1.5em)$) {};
    \node[myarrow] (p1-arr) at ($(p1) + (0em,1.5em)$) {};
    % Arrow labels
    \node[font=\scriptsize\sc,color=white] at (p0-arr) {nn2svgp};
    \node[font=\scriptsize\sc,color=white] at (p1-arr) {new data};   
  \end{tikzpicture}
\end{figure}

@@

\textbf{Neural Network} A neural network takes ${\mathcal{D}_n = \{(\vx_i, y_i)\}_{i=1}^n}$ and fits a function $f(\mathbf{x}_{i};\transitionParams)$ through the following optimization:
\begin{align} \label{eq-empirical-risk}
  \transitionParams^{*} = \arg \min_{\transitionParams \in \R^{P}} \mathcal{L}(\dataset;\transitionParams) = \arg \min_{\transitionParams \in \R^{P}} \left(
  \sum_{i=1}^{N} l(f(\mathbf{x}_{i} ; \transitionParams), y_i) + \frac{1}{\delta}R(\transitionParams)  \right).
\end{align}
If $l(f(\mathbf{x}_{i} ; \transitionParams), y_i) = -\log(p(y \mid f(\mathbf{x}_{i}))$

\textbf{Laplace approximation}

\textbf{Dual Gaussian Process} A Gaussian process (GP) is a prior over functions $f \sim \GP(0, \kappa)$. The prior is characterized by a covariance kernel function $\kappa(\vx,\vx')$, where $\vx, \vx'$ are input vectors typically in $\R^{\mathrm{D}}$. Once combined with a data set ${\mathcal{D}_n = \{(\vx_i, y_i)\}_{i=1}^n}$ with $n$ examples, each modeled by the likelihood $p(y_i \mid f_i)$ at function $f_i = f(\vx_i)$, we can form a posterior proccess using multivariate conditioning. The drawbacks to GPs are their lack of scaling to large datasets and non-tractability with non-Gaussian likelihoods.

Sparse Gaussian processes \citep{hensman2013gaussian} are a principled way to scale Gaussian processes to large datasets by changing the complexity from $O(n^3)$ to $O(m^3)$ with $m$ the number of inducing points, and typically $m \ll n$. The key insight is to use a sparse approximating distribution $q(\vu; \vm, \MV)$, where $\vm \in \R^{m}$ and $\MV \in \R^{m \times m}$, and find its parameters through maximizing the evidence lower bound (ELBO). The distribution at all over points is then defined as $q_{\vu}(f) = \int p(f \mid \vu) q(\vu) d\vu$. The dual parameterization \cite{adam2021dual} states that we can decompose the optimal sparse Gaussian process posterior as follows:
\begin{equation}
   \myexpect_{q_{\vu}(f)}[f_i] = \vkzs^{\T} \valpha , \quad 
   \textrm{Var}_{q_{\vu}(f)}[f_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta)^{-1} ]\vkzs.
\end{equation}

Here, $\MKzz$ to denote the $m \times m$ matrix with $\kappa(\vx_i,\vx_j)$ as the $ij$\textsuperscript{th} entry, Similarly, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $\kappa_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters consist of an $m$-length vector $\valpha$ and $m\times m$-size matrix $\vbeta$, defined as
\begin{equation} 
   \vlambda = [ \MKzz + \vbeta]\valpha  = \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \tilde{\lambda}_{i}, \quad
   \vbeta = \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,
\label{eq:dual_sparse}
\end{equation}

where $\tilde{\lambda}_i = \alpha_i - \myexpect_{q_{\vu}(f)}[f_i] \hat{\beta}_i$ with $(\hat{\alpha}_i, \hat{\beta}_i)$ defined as,
\begin{equation}
 \alpha_i = \myexpect_{q_{\vu}(f_i)} [\nabla_{f_i}\log p(y_i \mid f_i)], \quad
   \beta_i = \myexpect_{q_{\vu}(f_i)} [ -\nabla^{2}_{f_i}\log p(y_i \mid f_i)],
\label{eq:svgp_dual}
\end{equation}

>>>>

% Figure customization: Tight legend box
\pgfplotsset{every axis/.append style={
		legend style={inner xsep=1pt, inner ysep=0.5pt, nodes={inner sep=1pt, text depth=0.1em},draw=none,fill=none}
}}

@@

\newcommand{\vm}{\mbf{m}}
%\newcommand{\vz}{\mbf{z}}
\newcommand{\vf}{\mbf{f}}
\newcommand{\vu}{\mbf{u}}
\newcommand{\vx}{\mbf{x}}
\newcommand{\vy}{\mbf{y}}
\newcommand{\vw}{\mbf{w}}

@@

\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\Norm}{\mathcal{N}}

@@

\section{Introduction}
\label{sec:intro}
%
Deep learning has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations when applied to sequential learning paradigms, such as Continual Learning (CL, \cite{add}) and Reinforcement Learning (RL, \cite{add}). In CL, the challenge is retaining a compact representation of the problem and avoid forgetting over the life-long learning horizon (see, \eg, \cite{add}). Similarly in RL, the model would need to adapt to observations of the environment through exploration, but this can be inefficient and would require constant retraining (\eg, \cite{add, add}).

@@

\subsection{Related Work}
\label{sec:related}
%
% Introduce the laplace GNN + shortcomings
One common approach for uncertainty with neural networks is a Laplace GNN approximation \citep{daxbergerLaplace2021}. The idea is to essentially takes a trained neural network and linearises around the optimal weights. The resulting linear model can be used to obtain uncertainty estimates and sometimes improved predictions \citep{immerScalable2021}. Given the resultant model is now linear with respect to the weights its is also now possible to convert to a Gaussian process as shown in \citep{immerScalable2021, khan2019approximate}. Yet to obtain uncertainty estimates there is the usual computational scaling of the GP of $O(n^3)$, so cubic in the number of data points. Many approaches  apply crude approximations and yet still obtain good uncertainty estimates.

@@

\begin{figure}[t!]
  \centering\scriptsize
  % Figure options
  \pgfplotsset{axis on top,scale only axis,width=\figurewidth,height=\figureheight, ylabel near ticks,ylabel style={yshift=-2pt},y tick label style={rotate=90},legend style={nodes={scale=1., transform shape}},tick label style={font=\tiny,scale=1}}
  \pgfplotsset{xlabel={Input, $x$},axis line style={rounded corners=3pt}}
  % Set figure 
  \setlength{\figurewidth}{.28\textwidth}
  \setlength{\figureheight}{\figurewidth}
  %
  \def\inducing{\normalsize Sparse inducing points}
  %
  \begin{subfigure}[c]{.34\textwidth}
    \raggedleft
    \pgfplotsset{ylabel={Output, $y$}}
    \input{./fig/regression-nn.tex}%
  \end{subfigure}
  \hfill  
  \begin{subfigure}[c]{.01\textwidth}
    \centering
    \tikz[overlay,remember picture]\node(p0){};
  \end{subfigure}  
  \hfill
  \begin{subfigure}[c]{.28\textwidth}
    \raggedleft
    \pgfplotsset{yticklabels={,,},ytick={\empty}}
    \input{./fig/regression-nn2svgp.tex}%
  \end{subfigure}
  \hfill  
  \begin{subfigure}[c]{.01\textwidth}
    \centering
    \tikz[overlay,remember picture]\node(p1){};
  \end{subfigure}  
  \hfill
  \begin{subfigure}[c]{.28\textwidth}
    \raggedleft
    \pgfplotsset{yticklabels={,,},ytick={\empty}}        
    \input{./fig/regression-update.tex}%
  \end{subfigure}
  \caption{\textbf{Regression example on an MLP with two hidden layers.} Left:~Predictions from the trained neural network. Center:~Our approach summarizes the...}
  \label{fig:teaser} 
  % 
  \begin{tikzpicture}[remember picture,overlay]
    % Arrow style
    \tikzstyle{myarrow} = [draw=black!80, single arrow, minimum height=14mm, minimum width=2pt, single arrow head extend=4pt, fill=black!80, anchor=center, rotate=0, inner sep=5pt, rounded corners=1pt]
    % Arrows
    \node[myarrow] (p0-arr) at ($(p0) + (1em,1.5em)$) {};
    \node[myarrow] (p1-arr) at ($(p1) + (1em,1.5em)$) {};
    % Arrow labels
    \node[font=\scriptsize\sc,color=white] at (p0-arr) {nn2svgp};
    \node[font=\scriptsize\sc,color=white] at (p1-arr) {new data};   
  \end{tikzpicture}
\end{figure}

@@

\textbf{Neural Network} In supervised learning, given a data set $\dataset = \{(\mathbf{x}_{i} , \mathbf{y}_{i})\}_{i=1}^{N}$, with $\mathbf{x}_i \in \inputDomain$ and $\mathbf{y}_i \in \outputDomain$, the weights $\transitionParams \in \R^{P}$ of a neural network
, $f_\mathbf{w} : \inputDomain \mapsto \outputDomain$, 
are usually trained to minimize the (regularized) empirical risk,
\begin{align} \label{eq-empirical-risk}
  \transitionParams^{*} = 
  \argmin_{\transitionParams} \mathcal{L}(\dataset,\transitionParams) =
  \argmin_{\transitionParams} \sum_{i=1}^{N} \ell(f_\transitionParams(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\transitionParams).
\end{align}
If $\ell(f_\transitionParams(\mathbf{x}_{i}), \mathbf{y}_i) = -\log(p(\mathbf{y} \mid f_\transitionParams(\mathbf{x}_{i}))$ and $\mathcal{R}(\transitionParams) = \|\transitionParams\|^{2}_2$ then we can view \cref{eq-empirical-risk} as the MAP solution to a Bayesian objective, where $p(\vw) = \Norm(\vzeros, \delta^{-1} \MI)$.

\textbf{Laplace approximation} The Laplace approximation allows us to build a posterior approximation $q(\vw)$ from the optimized objective \cref{eq-empirical-risk}.

\textbf{Dual Gaussian Process} A Gaussian process (GP) is a prior over functions $f \sim \GP(\mathbf{0}, \kappa(\cdot, \cdot))$. The prior is characterized by a covariance kernel function $\kappa(\vx,\vx')$, where $\vx, \vx'$ are input vectors in $\R^{\mathrm{D}}$. Once combined with a data set ${\mathcal{D} = \{(\vx_i, y_i)\}_{i=1}^N}$ with $N$ examples, each modelled by the likelihood $p(\vy_i \mid \vf_i)$ at function $\vf_i = f(\vx_i)$\, we can form a posterior proccess using multivariate conditioning. The drawbacks to GPs are their lack of scaling to large datasets and non-tractability with non-Gaussian likelihoods.

Sparse Gaussian processes \citep{hensman2013gaussian} are a principled way to scale Gaussian processes to large datasets by using a subset of $M$ inducing points, with $M \ll N$,  lowering the complexity from $\mathcal{O}(N^3)$ to $\mathcal{O}(M^3)$. The key insight is to use a sparse approximating distribution $q(\vu; \vm, \MV)$, where $\vm \in \R^{M}$ and $\MV \in \R^{M \times M}$, and find its parameters through maximizing the evidence lower bound (ELBO). Then, the distribution over all points is defined as $q_{\vu}(\vf) = \int p(\vf \mid \vu) q(\vu) \mathrm{d}\vu$. As shown in \cite{adam2021dual}, we can decompose the optimal sparse GP  with a dual parameterization \cite{adam2021dual} as follows:
\begin{equation}
   \myexpect_{q_{\vu}(\vf)}[\vf_i] = \vkzs^{\T} \valpha , \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[\vf_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta)^{-1} ]\vkzs
\end{equation}

where the $ij$th entry of the matrix $\MKzz \in \R^{M \times M}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$\todo{the equation above uses $\vkzs$ not $\vk_{\vx i}$} denotes a vector where each $j$th element is $\kappa(\vx_i, \vx_j)$, and $\kappa_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters consist of the vector $\valpha \in \R^{M}$ and the matrix $\vbeta \in \R^{M\times M}$, defined as
\begin{equation} 
	\valpha  = \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \tilde{\lambda}_{i}, \quad
	\vbeta = \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    \vlambda = [ \MKzz + \vbeta]
\label{eq:dual_sparse}
\end{equation}

where $\tilde{\lambda}_i = \alpha_i - \myexpect_{q_{\vu}(\vf)}[\vf_i] \hat{\beta}_i$ with $(\hat{\alpha}_i, \hat{\beta}_i)$ defined as,
\begin{equation}
 \alpha_i = \myexpect_{q_{\vu}(\vf_i)} [\nabla_{\vf_i}\log p(\vy_i \mid \vf_i)], \quad
   \beta_i = \myexpect_{q_{\vu}(\vf_i)} [ -\nabla^{2}_{\vf_i}\log p(\vy_i \mid \vf_i)],
\label{eq:svgp_dual}
\end{equation}
